How to select predictive models for causal inference?

### 2023-11-16 how_to_select_causal_models.tex
finish to adpat paper

%\idea{There is a variety of models for g-estimation}
There is an explosion of outcome modeling or machine learning methods.
For instance, many deep-learning methods have been developed for medical
image analysis \cite{shen2017deep,monshi2020deep}.
Even outcome-modeling methods specifically designed for causal
inference are numerous: Bayesian Additive Regression Trees
\cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
\cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
\cite{powers_methods_2018}, causal multivariate adaptive regression
splines \cite{powers_methods_2018}, random forests
\cite{wager_estimation_2018, athey_generalized_2019},
Meta-learners \cite{kunzel_metalearners_2019}, R-learners
\cite{nie_quasioracle_2017}, Doubly robust estimation
\cite{chernozhukov_double_2018}...
%\idea{The variety of estimators calls for model selection procedures}
The wide variety of methods raises the problem
of selecting between different estimators based on the data at hand.
%
Indeed, estimates of treatment effects can vary markedly across different
predictive models. For instance, Figure shows
large variations obtained across different outcome estimators on
semi-synthetic datasets \cite{dorie_automated_2019}. Flexible models
such as random forests are doing well in most settings except
when treated and untreated populations differ noticeably, in
which case a linear model (ridge) is to be preferred.
However random forests with different hyper-parameters
(max depth= 2) yield poor estimates.
A simple rule of thumb such as preferring flexible models does not work in
general; model selection is needed.

@@

We first illustrate the problem of causal model
selection and briefly review prior art. Then, Section
\ref{sec:framework} sets causal model selection in the
\emph{potential outcome} framework and details the causal risks and
model-selection procedure. Section \ref{sec:theory} gives theoretical
results. Section \ref{sec:empirical_study} details a thorough empirical
study, covering many different settings. Finally, Section \ref{sec:discussion}
discusses the findings.
%
Results outline how to best select outcome models for causal
inference with an adapted
cross-validation to estimate the so-called $R\text{-risk}$.
This risk compensates for systematic
differences between treated and non-treated individuals using
two \emph{nuisance} models,
themselves estimated from data and thus imperfect; yet these
imperfections do not undermine the $R\text{-risk}$.

@@

\begin{figure}[t!]
    %\centering
    {\sffamily\footnotesize\scalebox{1.03}{a) Random forest, good average prediction but bad
            causal inference}}

@@

\begin{table*}[!tb]
    \makebox[\textwidth]{
        \begin{threeparttable}[b]
            \caption{Review of causal risks
                ---
                The $R\text{-risk}^*$ is called $\tau \text{-risk}_R$ in
                \citet{schuler_comparison_2018}.
                \label{tab:evaluation_metrics}}
            \centering
            \begin{tabular}{llr}
                \toprule
                Risk                                                                                           & Equation
                                                                                                               & Reference                                                                                                                                           \\
                \midrule
                $mse(\tau(X), \tau_f(X))=\tau\text{-risk}$                                                     & $\mathbb E_{X\sim
                            p(X)}[(\tau(X) - \hat \tau_f(X))^2] $
                                                                                                               & Eq. \ref{eq:tau_risk} \cite{hill_bayesian_2011}                                                                                                     \\
                $mse(Y, f(X)) = \mu\text{-risk}$                                                               & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D}\left[(Y-f(X ; A))^2 \right]$
                                                                                                               & Def. \ref{def:mu_risk} \cite{schuler_comparison_2018}                                                                                               \\
                $\mu\text{-risk}_{IPW}^*$                                                                      & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D}\left[ \Big( \frac{A}{e(X)} + \frac{1-A}{1-e(X)} \Big)
                (Y-f(X ; A))^2 \right]$                                                                        & Def.
                \ref{def:mu_ipw_risk} \cite{vanderlaan_unified_2003}                                                                                                                                                                                                 \\
                $\tau\text{-risk}^{\star}_{IPW}$                                                               & $\mathbb{E}_{(Y, X, A) \sim \mathcal D} \left[ \Big(Y \left( \frac{A}{e(X)} - \frac{1-A}{1-e(X)}\right)-\hat \tau_f\left(X\right)\Big)^2 \right]$ &
                Def. \ref{def:tau_ipw_risk} \cite{wager_estimation_2018}
                \\
                $U\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A) \sim \mathcal D}  \big[
                \big( \frac{Y-m\left(X\right)}{A-e\left(X\right)} -  \hat \tau_f\left(X\right)\big)^{2} \big]$ &
                Def. \ref{def:u_risk} \cite{nie_quasioracle_2017}
                \\
                $R\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D} \big[\big(\left(Y-m\left(X\right)\right)
                -\left(A-e\left(X\right)\right) \hat \tau_f\left(X\right)\big)^{2} \big]$                      &
                Def. \ref{def:r_risk} \cite{nie_quasioracle_2017}
                \\
                \bottomrule
            \end{tabular}
        \end{threeparttable}
    }
\end{table*}

@@

\paragraph{Randomization special case}\label{remark:rct} If the treatment is
randomized as in RCTs, $p(A=1 \mid X=x) = p(A=1)=p_A$, thus
$\mu\text{-risk}_{IPW}$ takes a simpler form:
\begin{equation*}
    \mu\text{-risk}_{IPW} = \mathbb{E}_{(Y, X, A) \sim \mathcal D}\left[ \Big( \frac{A}{p_A} + \frac{1-A}{1-p_A} \Big) (Y-f(X ; A))^2 \right]
\end{equation*}
However, we still can have large differences
between $\tau\text{-risk}$ and $\mu\text{-risk}_{IPW}$ coming from heterogeneous
errors between populations as noted in Section \ref{theory:mu_risk_ipw_bound}
and shown experimentally in \citet{schuler_comparison_2018} and our
results below.

@@

The number of basis functions --\emph{ie. knots}--, controls the complexity of
the ground-truth response surfaces and treatment. We first use this process to
draw the non-treated response surface $\mu_0$ and the causal effect $\tau$. We
then draw the observations from a mixture two Gaussians, for the treated and non
treated. We vary the separation between the two Gaussians to control the
overlap between treated and non-treated populations, an important parameter
for causal inference (related to $\eta$ in section
\ref{theory:mu_risk_ipw_bound}). Finally, we generate observed outcomes
adding Gaussian noise. We generate 1\,000 of such datasets, with
uniformly random overlap parameters. Details in
\ref{apd:experiments:generation}.

@@

Predictive models are increasingly used to reason about causal effects,
for instance in precision medicine to drive individualized decision.
Our results highlight that they should be selected, validated, and tuned
using different procedures and error measures than those classically used
to assess prediction (estimating the so-called $\mu\text{-risk}$).
Rather, selecting the best outcome model according to the $R\text{-risk}$
(eq.\,\autoref{def:r_risk}) leads to more valid causal estimates.

@@

The nuisance models must be well estimated (Figure
\ref{fig:all_datasets_nuisances_comparison}). However these models are
easier to select and control than a causally-valid outcome model,
as they are associated to errors on
observed distributions. Our results show that using for nuisance models
a flexible stacking-based family of estimator suffices for good model selection.
%
In fact, a feasible $R\text{-risk}$ --where the nuisances
are estimated-- performs almost as well as an oracle $R\text{-risk}$ --where the
nuisances are known. This may be explained by
results that suggest that estimation errors on both
nuisances partly compensate out in the
$R\text{-risk} $\cite{daniel2018double,kennedy2020optimal,nie_quasioracle_2017,chernozhukov_double_2018,zivich2021machine,naimi2021challenges}.

@@

\paragraph{Motivation of the Normalized Total Variation}
%\idea{A simplier measure, NTV seems to capture what we want}
Computing overlap when working only on samples of the observed distribution,
outside of simulation, requires a sophisticated estimator of discrepancy
between distributions, as two data points never have the same exact set of
features. Maximum Mean Discrepancy \cite{gretton2012kernel} is typically
used in the context of causal inference
\cite{shalit_estimating_2017,johansson_generalization_2021}. However it
needs a kernel, typically Gaussian, to extrapolate across neighboring
observations. We prefer avoiding the need to specify such a kernel, as it must
be adapted to the data which is tricky with categorical or non-Gaussian
features, a common situation for medical data.

@@

@@

Before even model selection, data splitting is often required for
estimation of the treatment effect, ATE or CATE, for instance to compute
the nuisances required to optimize the outcome model (as the
$R\text{-risk}$, definition \autoref{def:r_risk}).
%
The most frequent choice is use 80\% of the data to fit the models,
and 20\% to evaluate them.
%
For instance, for CATE estimation, the R-learner has been introduced using K-folds with K = 5
and K = 10: 80\% of the data (4 folds) to train the nuisances and the remaining
fold to minimize the corresponding R-loss \cite{nie_quasioracle_2017}.
Yet, it has been implemented with K=5 in causallib
\cite{causalevaluations} or K=3 in econML \cite{econml}.
Likewise, for ATE estimation, \citet{chernozhukov_double_2018}
introduce doubly-robust machine learning,
recommending K=5 based on an empirical comparison K=2. However,
subsequent works use doubly robust ML with varying choices
of K: \citet{loiseau_external_2022} use K=3, \citet{gao_assessment_2021} use
K=2. In the econML implementation, K is set to 3 \cite{econml}.
\citet{naimi2021challenges} evaluate various machine-learning approaches
--including R-learners-- using K=5 and 10, drawing inspiration from the
TMLE literature which sets
K=5 in the TMLE package \cite{tmle_package_2012}.

>>>>

%\idea{There is a variety of models for g-estimation}
There is an explosion of outcome modeling or machine learning methods.
For instance, many deep-learning methods have been developed for medical
image analysis \cite{shen2017deep,monshi2020deep}.
Even outcome-modeling methods specifically designed for causal
inference are numerous: Bayesian Additive Regression Trees
\cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
\cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
\cite{powers_methods_2018}, causal multivariate adaptive regression
splines \cite{powers_methods_2018}, random forests
\cite{wager_estimation_2018, athey_generalized_2019},
Meta-learners \cite{kunzel_metalearners_2019}, R-learners
\cite{nie_quasioracle_2017}, Doubly robust estimation
\cite{chernozhukov_double_2018}...
%\idea{The variety of estimators calls for model selection procedures}
The wide variety of methods raises the problem
of selecting between different estimators based on the data at hand.
%
Indeed, estimates of treatment effects can vary markedly across different
predictive models. For instance, Figure \ref{fig:acic_2016_ate_heterogeneity} shows
large variations obtained across different outcome estimators on
semi-synthetic datasets \cite{dorie_automated_2019}. Flexible models
such as random forests are doing well in most settings except
when treated and untreated populations differ noticeably, in
which case a linear model (ridge) is to be preferred.
However random forests with different hyper-parameters
(max depth= 2) yield poor estimates.
A simple rule of thumb such as preferring flexible models does not work in
general; model selection is needed.

@@

We first illustrate the problem of causal model
selection and briefly review prior art. Then the methods
\autoref{sec:framework} sets causal model selection in the
\emph{potential outcome} framework and details the causal risks and
model-selection procedure. The next \autoref{sec:theory} gives theoretical
results. The results \autoref{sec:empirical_study} details a thorough empirical
study, covering many different settings. Finally, we
discuss the findings in the last \autoref{sec:discussion}.
%
Results outline how to best select outcome models for causal
inference with an adapted
cross-validation to estimate the so-called $R\text{-risk}$.
This risk compensates for systematic
differences between treated and non-treated individuals using
two \emph{nuisance} models,
themselves estimated from data and thus imperfect; yet these
imperfections do not undermine the $R\text{-risk}$.

@@

\begin{figure}[b!]
    %\centering
    {\sffamily\footnotesize\scalebox{1.03}{a) Random forest, good average prediction but bad
            causal inference}}

@@

\begin{table*}[bt!]
    \makebox[\textwidth]{
        \begin{threeparttable}[b]
            \caption{Review of causal risks
                ---
                The $R\text{-risk}^*$ is called $\tau \text{-risk}_R$ in
                \citet{schuler_comparison_2018}.
                \label{tab:evaluation_metrics}}
            \centering
            \begin{tabular}{llr}
                \toprule
                Risk                                                                                           & Equation
                                                                                                               & Reference                                                                                                                                           \\
                \midrule
                $mse(\tau(X), \tau_f(X))=\tau\text{-risk}$                                                     & $\mathbb E_{X\sim
                            p(X)}[(\tau(X) - \hat \tau_f(X))^2] $
                                                                                                               & Eq. \ref{eq:tau_risk} \cite{hill_bayesian_2011}                                                                                                     \\
                $mse(Y, f(X)) = \mu\text{-risk}$                                                               & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D}\left[(Y-f(X ; A))^2 \right]$
                                                                                                               & Def. \ref{def:mu_risk} \cite{schuler_comparison_2018}                                                                                               \\
                $\mu\text{-risk}_{IPW}^*$                                                                      & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D}\left[ \Big( \frac{A}{e(X)} + \frac{1-A}{1-e(X)} \Big)
                (Y-f(X ; A))^2 \right]$                                                                        & Def.
                \ref{def:mu_ipw_risk} \cite{vanderlaan_unified_2003}                                                                                                                                                                                                 \\
                $\tau\text{-risk}^{\star}_{IPW}$                                                               & $\mathbb{E}_{(Y, X, A) \sim \mathcal D} \left[ \Big(Y \left( \frac{A}{e(X)} - \frac{1-A}{1-e(X)}\right)-\hat \tau_f\left(X\right)\Big)^2 \right]$ &
                Def. \ref{def:tau_ipw_risk} \cite{wager_estimation_2018}
                \\
                $U\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A) \sim \mathcal D}  \big[
                \big( \frac{Y-m\left(X\right)}{A-e\left(X\right)} -  \hat \tau_f\left(X\right)\big)^{2} \big]$ &
                Def. \ref{def:u_risk} \cite{nie_quasioracle_2017}
                \\
                $R\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D} \big[\big(\left(Y-m\left(X\right)\right)
                -\left(A-e\left(X\right)\right) \hat \tau_f\left(X\right)\big)^{2} \big]$                      &
                Def. \ref{def:r_risk} \cite{nie_quasioracle_2017}
                \\
                \bottomrule
            \end{tabular}
        \end{threeparttable}
    }
\end{table*}

@@

\paragraph{Randomization special case}\label{remark:rct} If the treatment is
randomized as in RCTs, $p(A=1 \mid X=x) = p(A=1)=p_A$, thus
$\mu\text{-risk}_{IPW}$ takes a simpler form:
\begin{equation*}
    \mu\text{-risk}_{IPW} = \mathbb{E}_{(Y, X, A) \sim \mathcal D}\left[ \Big( \frac{A}{p_A} + \frac{1-A}{1-p_A} \Big) (Y-f(X ; A))^2 \right]
\end{equation*}
However, we still can have large differences
between $\tau\text{-risk}$ and $\mu\text{-risk}_{IPW}$ coming from heterogeneous
errors between populations as noted by Proposition \ref{theory:prop:mu_risk_ipw_bound}
and shown experimentally in \citet{schuler_comparison_2018} and our
results below.

@@

The number of basis functions --\emph{ie. knots}--, controls the complexity of
the ground-truth response surfaces and treatment. We first use this process to
draw the non-treated response surface $\mu_0$ and the causal effect $\tau$. We
then draw the observations from a mixture two Gaussians, for the treated and non
treated. We vary the separation between the two Gaussians to control the
overlap between treated and non-treated populations, an important parameter
for causal inference (related to $\eta$ in Proposition
\ref{theory:prop:mu_risk_ipw_bound}). Finally, we generate observed outcomes
adding Gaussian noise. We generate 1\,000 of such datasets, with
uniformly random overlap parameters. Example shown in Figure \ref{fig:simulation_examples} and details in
\ref{apd:experiments:generation}.

@@

Predictive models are increasingly used to reason about causal effects,
for instance in precision medicine to drive individualized decision.
Our results highlight that they should be selected, validated, and tuned
using different procedures and error measures than those classically used
to assess prediction (estimating the so-called $\mu\text{-risk}$).
Rather, selecting the best outcome model according to the $R\text{-risk}$
(eq.\, Definition \ref{def:r_risk}) leads to more valid causal estimates.

@@

The nuisance models must be well estimated (Figure
\ref{fig:all_datasets_nuisances_comparison}). However these models are
easier to select and control than a causally-valid outcome model,
as they are associated to errors on
observed distributions. Our results show that using for nuisance models
a flexible stacking-based family of estimator suffices for good model selection.
%
In fact, a feasible $R\text{-risk}$ --where the nuisances
are estimated-- performs almost as well as an oracle $R\text{-risk}$ --where the
nuisances are known. This may be explained by
results that suggest that estimation errors on both
nuisances partly compensate out in the
$R\text{-risk}$ \cite{daniel2018double,kennedy2020optimal,nie_quasioracle_2017,chernozhukov_double_2018,zivich2021machine,naimi2021challenges}.

@@

\paragraph{Motivation of the Normalized Total Variation}
%\idea{A simplier measure, NTV seems to capture what we want}
Computing overlap when working only on samples of the observed distribution,
outside of simulation, requires a sophisticated estimator of discrepancy
between distributions, as two data points never have the same exact set of
features. Maximum Mean Discrepancy \cite{gretton2012kernel} is typically
used in the context of causal inference
\cite{shalit_estimating_2017,johansson2022generalization}. However it
needs a kernel, typically Gaussian, to extrapolate across neighboring
observations. We prefer avoiding the need to specify such a kernel, as it must
be adapted to the data which is tricky with categorical or non-Gaussian
features, a common situation for medical data.

@@

\paragraph{Figure \ref{apd:fig:relative_kendalls_all_datasets_all_metrics} -
    Results measured in relative Kendall's for feasible and semi-oracle risks}

Because of extreme propensity scores in the denominator and bayes error residuals in the numerator, the semi-oracle
$U$-risk has poor performances at bad overlap. Estimating these propensity scores in the is feasible $U$-risk reduces
the variance since clipping is performed.

\begin{figure}[!b]
    \centering
    \includegraphics[width=\linewidth]{images/_1_r_risk_domination_r_risk_domination__ref_metric_mean_risks_by_Dataset.pdf}
    %\hfill
    \caption{\textbf{The $R$-risk is the best metric}: Relative Kendall's $\tau$
        agreement with $\tau\text{-risk}$. Strong and Weak overlap correspond to
        the first and last tertiles of the overlap distribution measured with
        Normalized Total Variation eq. \ref{eq:ntv}.
    }\label{apd:fig:relative_kendalls_all_datasets_all_metrics}
\end{figure}

@@

Before even model selection, data splitting is often required for
estimation of the treatment effect, ATE or CATE, for instance to compute
the nuisances required to optimize the outcome model (as the
$R\text{-risk}$, Definition \ref{def:r_risk}).
%
The most frequent choice is use 80\% of the data to fit the models,
and 20\% to evaluate them.
%
For instance, for CATE estimation, the R-learner has been introduced using K-folds with K = 5
and K = 10: 80\% of the data (4 folds) to train the nuisances and the remaining
fold to minimize the corresponding R-loss \cite{nie_quasioracle_2017}.
Yet, it has been implemented with K=5 in causallib
\cite{causalevaluations} or K=3 in econML \cite{econml}.
Likewise, for ATE estimation, \citet{chernozhukov_double_2018}
introduce doubly-robust machine learning,
recommending K=5 based on an empirical comparison K=2. However,
subsequent works use doubly robust ML with varying choices
of K: \citet{loiseau_external_2022} use K=3, \citet{gao_assessment_2021} use
K=2. In the econML implementation, K is set to 3 \cite{econml}.
\citet{naimi2021challenges} evaluate various machine-learning approaches
--including R-learners-- using K=5 and 10, drawing inspiration from the
TMLE literature which sets
K=5 in the TMLE package \cite{tmle_package_2012}.
